---
title: "Marine Proxies"
author: "Gabriel Bowen"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
editor: source
format: pdf
---

```{r setup, include=FALSE}
source("code/prepData.R")
load("bigout/post.rda")
source("code/prepBridge.R")

pm = post$BUGSoutput$median
sl = post$BUGSoutput$sims.list

col.c = rgb(1, 0.2, 0.6)
col.a = rgb(0.2, 0.6, 1)

knitr::opts_knit$set(global.par = TRUE)
```

Let's look at the marine proxy data. The latest run includes ocean average $\delta^{18}O$ value as a free parameter and the high-latitude continental data for both stages.

```{r gradient}
#| echo: FALSE
#| fig.height: 10
l = 0:90
layout(matrix(c(1, 2), ncol = 1))
par(mar = c(5, 5, 1, 1))

plot(l, 23 + l * 0.22 + l^2 * -0.008, type = "l", lwd = 2, ylim = c(0,43),
     xlim = c(2, 80), xlab = "Latitude", ylab = "Mean temperature")
for(i in seq(1, nrow(sl$a.car), length.out = 1000)){
  lines(l, sl$a.car[i] + l * 
          sl$b.car[i] + 
          l^2 * sl$c.car[i], 
        col = rgb(1, 0.2, 0.6, 0.02))
}
lines(l, c(post$BUGSoutput$mean$a.car) + l * c(post$BUGSoutput$mean$b.car) + 
        l^2 * c(post$BUGSoutput$mean$c.car), col = col.c, lwd = 2)

lines(zMeans$lat, zMeans$car, col = col.c)
lines(zMeans$lat, zMeans$carC, col = col.c, lty = 2)

arrows(post$BUGSoutput$mean$lat_mar.car, 
       post$BUGSoutput$mean$t_mar.car - post$BUGSoutput$sd$t_mar.car, 
       post$BUGSoutput$mean$lat_mar.car, 
       post$BUGSoutput$mean$t_mar.car + post$BUGSoutput$sd$t_mar.car, 
       angle = 90, length = 0.05, code = 3, col = col.c)
points(post$BUGSoutput$mean$lat_mar.car, 
       post$BUGSoutput$mean$t_mar.car, pch = 21, bg = "white", 
       col = col.c)
arrows(post$BUGSoutput$mean$lat_cont.car, 
       post$BUGSoutput$mean$t_cont.car - post$BUGSoutput$sd$t_cont.car, 
       post$BUGSoutput$mean$lat_cont.car, 
       post$BUGSoutput$mean$t_cont.car + post$BUGSoutput$sd$t_cont.car, 
       angle = 90, length = 0.05, code = 3, 
       col = col.c)
points(post$BUGSoutput$mean$lat_cont.car, post$BUGSoutput$mean$t_cont.car, 
       pch = 21, bg = "dark grey", col = col.c)

plot(l, 23 + l * 0.22 + l^2 * -0.008, type = "l", lwd = 2, ylim = c(0,43),
     xlim = c(2, 80), xlab = "Latitude", ylab = "Mean temperature")

for(i in seq(1, nrow(sl$a.an), length.out = 1000)){
  lines(l, sl$a.an[i] + l * 
          sl$b.an[i] + 
          l^2 * sl$c.an[i], 
        col = rgb(0.2, 0.6, 1, 0.02))
}
lines(l, c(post$BUGSoutput$mean$a.an) + l * c(post$BUGSoutput$mean$b.an) + 
        l^2 * c(post$BUGSoutput$mean$c.an), col = col.a, lwd = 2)

lines(zMeans$lat, zMeans$an, col = col.a)
lines(zMeans$lat, zMeans$anC, col = col.a, lty = 2)

arrows(post$BUGSoutput$mean$lat_mar.an, 
       post$BUGSoutput$mean$t_mar.an - post$BUGSoutput$sd$t_mar.an, 
       post$BUGSoutput$mean$lat_mar.an, 
       post$BUGSoutput$mean$t_mar.an + post$BUGSoutput$sd$t_mar.an, 
       angle = 90, length = 0.05, code = 3, col = col.a)
points(post$BUGSoutput$mean$lat_mar.an, 
       post$BUGSoutput$mean$t_mar.an, pch = 21, bg = "white", 
       col = col.a)
arrows(post$BUGSoutput$mean$lat_cont.an, 
       post$BUGSoutput$mean$t_cont.an - post$BUGSoutput$sd$t_cont.an, 
       post$BUGSoutput$mean$lat_cont.an, 
       post$BUGSoutput$mean$t_cont.an + post$BUGSoutput$sd$t_cont.an, 
       angle = 90, length = 0.05, code = 3, 
       col = col.a)
points(post$BUGSoutput$mean$lat_cont.an, post$BUGSoutput$mean$t_cont.an, 
       pch = 21, bg = "dark grey", col = col.a)
```

Let's review the high-latitude (terrestrial) proxy data and its interpretation. I'll refer to the sites by number for convenience, here's the key:

```{r sites, echo=FALSE}
print("Carnian = pink")
mar_sites.car

print("Anisian = blue")
mar_sites.an
```

Here are the proxy data for marine calcite: 

```{r setpar}
#| include: false
dev.off()
par(mar = c(5, 5, 4, 1))
```

```{r calcite,fig.height=4}
#| echo: false
MAT = 15:45
plot(MAT, (0 + (2.78e6 / (MAT + 273) ^ 2 + -2.98)) * 0.97002 - 29.98, 
     xlim = c(18, 40), 
     ylim = range(c(mc_samples.car$d18O, mc_samples.car$d18O)), type = "l", 
     xlab = "MAT", ylab = expression(delta^{18}*"O marine calcite"))
lines(MAT, (c(pm$dO_mar.mu) + (2.78e6 / (MAT + 273) ^ 2 + -2.98)) * 0.97002 - 29.98,
      lty = 3)
points(pm$t_mar.car[mc_sites.ind.car], mc_samples.car$d18O, pch = 20, col = col.c)
points(pm$t_mar.an[mc_sites.ind.an], mc_samples.an$d18O, pch = 20, col = col.a)

text(pm$t_mar.car[mc_sites.ind.car], par("usr")[4], mc_sites.ind.car, 
     xpd = NA, pos = 3, col = col.c)

text(pm$t_mar.an[mc_sites.ind.an], par("usr")[4], mc_sites.ind.an, 
     xpd = NA, pos = 3, offset = 1.25, col = col.a)
```

The solid and dashed lines represent the paleotemperature equation using a seawater $\delta^{18}O$ value of zero (solid) or -1.1 per mil (dashed; this is the median value from the JPI posterior). A few observations:

* Comparing the data with the solid line, you can see why the tropics were so hot in our previous analyses: the calcite $\delta^{18}O$ values are very low. Aside from a small number of Carnian samples (pink) that plot nicely on the solid line, most of the data fall well below it. In the analyses that used the fixed seawater value of zero per mil the temperatures at these sites were pulled stongly toward higher values.
* Allowing for lower seawater $\delta^{18}O$ values makes lower MAT values more consistent with these low carbonate $\delta^{18}O$ values. Because most of the continental proxy data are consistent with pretty low MAT (as described in the previous analysis), and because the parameterization of the meridional MAT gradient *prefers* gradients that are not so steep, the tropical marine temperaures decline substantially when we allow this change.
* The distributions of carbonate values for the Carnian and Anisian are offset pretty strongly. I am assuming seawater $\delta^{18}O$ values are the same during both time periods (which I think is appropriate), so this translates to an offset in tropical MATs of ~10 degrees between the two stages.

For comparison here are the same plots for phosphates and dolomites.

```{r phosphate,fig.height=4}
#| echo: FALSE
MAT = 15:45
plot(MAT, (MAT - 118.7) / -4.22 + 0, xlim = c(18, 40),
     ylim = range(c(mp_samples.car$d18O, mp_samples.car$d18O)), type = "l", 
     xlab = "MAT", ylab = expression(delta^{18}*"O marine phosphate"))
lines(MAT, (MAT - 118.7) / -4.22 + c(pm$dO_mar.mu), lty = 3)
points(pm$t_mar.car[mp_sites.ind.car], mp_samples.car$d18O, pch = 20, col = col.c)
points(pm$t_mar.an[mp_sites.ind.an], mp_samples.an$d18O, pch = 20, col = col.a)


text(pm$t_mar.car[mp_sites.ind.car], par("usr")[4], mp_sites.ind.car, 
     xpd = NA, pos = 3, col = col.c)

text(pm$t_mar.an[mp_sites.ind.an], par("usr")[4], mp_sites.ind.an, 
     xpd = NA, pos = 3, offset = 1.25, col = col.a)
```

The MATs would again be much higher using a seawater $\delta^{18}O$ of zero. Interestingly, the phosphate $\delta^{18}O$ are much more similar for the Anisian and Carnian than the values for calcite are, and if we just used phosphate data the marine tropical temperatures for the two stages would be pretty similar. Because there are a lot more calcite data these end up being more influential, though.

```{r dolomite,fig.height=4}
#| echo: FALSE
MAT = 15:45
plot(MAT, (0 + (2.73e6 / (MAT + 273) ^ 2 + 0.26)) * 0.97002 - 29.98, 
     xlim = c(18, 40),
     ylim = range(md_samples.car$d18O), type = "l", xlab = "MAT", 
     ylab = expression(delta^{18}*"O marine dolomite"))
lines(MAT, (c(pm$dO_mar.mu) + (2.73e6 / (MAT + 273) ^ 2 + 0.26)) * 0.97002 - 29.98,
      lty = 3)
points(pm$t_mar.car[md_sites.ind.car], md_samples.car$d18O, pch = 20, col = col.c)


text(pm$t_mar.car[md_sites.ind.car], par("usr")[4], md_sites.ind.car, 
     xpd = NA, pos = 3, col = col.c)
```

Not so sure about these dolomite data...with the exception of a few low-value outliers they are consistent w/ substantially lower temperatures regardless of what we use for the $\delta^{18}O$ value of seawater. It's only one site, though and in the full analysis they are largley overwhelmed by the other data. 
