---
title: "Terrestrial Proxies"
author: "Gabriel Bowen"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
editor: source
---

```{r setup, include=FALSE}
source("code/prepData.R")
load("bigout/post.rda")

pm = post$BUGSoutput$median

col.c = rgb(1, 0.2, 0.6)
col.a = rgb(0.2, 0.6, 1)

knitr::opts_knit$set(global.par = TRUE)
```

```{r setpar, include=FALSE}
par(mar = c(5, 5, 4, 1))
```

Let's review the high-latitude (terrestrial) proxy data and its interpretation. I'll refer to the sites by number for convenience, here's the key:

```{r sites, echo=FALSE}
print("Carnian = pink")
cont_sites.car

print("Anisian = blue")
cont_sites.an
```


## Al/Si

Al/Si is available from 4 Carnian sites and 1 Anisian site. Here I'm plotting the calibration data set for the proxy (black) and the paleo-proxy values vs. the posterior median site MAT (pink = Carnian, blue = Anisian).

```{r alsi, echo=FALSE}
plot(AlSi ~ MAT, data = alsi_cal, ylab = "Al/Si",
     xlim = range(c(as.data.frame(alsi_cal)$MAT, pm$t_cont.car[alsi_sites.ind.car],
                    pm$t_cont.an[alsi_sites.ind.an])),
     ylim = range(c(as.data.frame(alsi_cal)$AlSi, alsi_samples.car$AlSi,
                    alsi_samples.an$AlSi)))

points(pm$t_cont.car[alsi_sites.ind.car], alsi_samples.car$AlSi,
       pch = 20, col = col.c)

points(pm$t_cont.an[alsi_sites.ind.an], alsi_samples.an$AlSi,
       pch = 20, col = col.a)

text(pm$t_cont.car[alsi_sites.ind.car], par("usr")[4], alsi_sites.ind.car, 
     xpd = NA, pos = 3, col = col.c)

text(pm$t_cont.an[alsi_sites.ind.an], par("usr")[4], alsi_sites.ind.an, 
     xpd = NA, pos = 3, offset = 1.25, col = col.a)
```

My observations:

* The data are very noisy.
* The posterior Carnian MATs skew a little bit cooler than you would get if you used these data alone, but are not too dissimilar on average.
* Carnian site 3 (Chanares) stands out as having reconstructed MAT substantially lower than the Al/Si would imply, alone.

## CIA

Same thing for CIA, which is available at 4 Carnian and 3 Anisian sites.

```{r CIA, echo=FALSE}
plot(CIA ~ Temp, data = cia_cal, xlab = "MAT",
     xlim = range(c(as.data.frame(cia_cal)$Temp, pm$t_cont.car[cia_sites.ind.car],
                    pm$t_cont.an[cia_sites.ind.an])),
     ylim = range(c(as.data.frame(cia_cal)$CIA, cia_samples.car$CIA,
                    cia_samples.an$CIA)))

points(pm$t_cont.car[cia_sites.ind.car], cia_samples.car$CIA,
       col = col.c, pch = 20)
points(pm$t_cont.an[cia_sites.ind.an], cia_samples.an$CIA,
       col = col.a, pch = 20)

text(pm$t_cont.car[cia_sites.ind.car], par("usr")[4], cia_sites.ind.car, 
     xpd = NA, pos = 3, col = col.c)
text(pm$t_cont.an[cia_sites.ind.an], par("usr")[4], cia_sites.ind.an, 
     xpd = NA, pos = 3, offset = 1.25, col = col.a)
```

* In this case, the posterior MATs for some sites tend to be a little higher than you'd expect just based on the proxy data (top/right).
* The MATs for the new Antarctic site plot just slightly cold compared with what you'd expect based on the CIA values.
* It's a really weak proxy (huge scatter in the calibration data) so not surprising that here, too, there's not a a very strong influence on the MATs.

## PWI

Now paleosol weathering index, available at one Anisian and two Carnian sites. I've added the calibration curve using the median curve fit parameters in the posterior for reference.

```{r PWI, echo=FALSE}
plot(PWI ~ MAT, data = pwi_cal, 
     xlim = range(c(as.data.frame(pwi_cal)$MAT, pm$t_cont.car[pwi_sites.ind.car],
                    pm$t_cont.an[pwi_sites.ind.an])),
     ylim = range(c(as.data.frame(pwi_cal)$PWI, pwi_samples.car$PWI, 
                    pwi_samples.an$PWI)))
xs = seq(par("usr")[1], par("usr")[2], length = 100)
ys = exp((xs + c(pm$pwi_a)) / c(pm$pwi_b))
lines(xs, ys)

points(pm$t_cont.car[pwi_sites.ind.car], pwi_samples.car$PWI,
       col = col.c, pch = 20)
points(pm$t_cont.an[pwi_sites.ind.an], pwi_samples.an$PWI,
       col = col.a, pch = 20)

text(pm$t_cont.car[pwi_sites.ind.car], par("usr")[4], pwi_sites.ind.car, 
     xpd = NA, pos = 3, col = col.c)
text(pm$t_cont.an[pwi_sites.ind.an], par("usr")[4], pwi_sites.ind.an, 
     xpd = NA, pos = 3, offset = 1.25, col = col.a)
```

Here the reconstructions fit the calibration data well. Obviously the interpretation is based on an extrapolation at site 3, but if one takes the proxy at face value it's pretty tough for this site to have a MAT higher than, say, 5 or 10 $^\circ$. Note that this is the same site which has a cool bias in the first plot (based on Al/Si).

The MAT for the new Antarctic site is slightly higher than expected based just on PWI (compare w/ the CIA result).

## Clumped

Here I'm using the calibration function from Julia Kelson's paper, which is represented as a line. Only one Anisian site.

```{r clumped, echo=FALSE}
plot(20:45, 6.36e4 / (20:45 + 273) ^ 2 + -4.7e-3, type = "l", 
     xlab = "MAT", ylab = expression(Delta*47),
     ylim = range(c(6.36e4 / (20:45 + 273) ^ 2 + -4.7e-3, 
                    clump_samples.an$cap_47)))

points(pm$t_cont.an[clump_sites.ind.an], clump_samples.an$cap_47,
       pch = 20, col = col.a)

points(pm$t_seas.an[clump_sites.ind.an], clump_samples.an$cap_47,
       col = col.a)

text(pm$t_cont.an[clump_sites.ind.an], par("usr")[4], clump_sites.ind.an, 
     xpd = NA, pos = 3, col = col.a)
```

* The JPI posterior is a little warm at this site for what you'd expect accounting for seasonality. The closed symbols are MAT, the open ones are summer temperature.
* Note that this same site has a bunch of CIA data shown above, and the posterior MAT is also a bit warmer than expected given those data. This is a sub-tropical site, and the fairly warm SSTs are pulling the MAT values at this and other sub-tropical Anisian sites toward higher values.

## Oxygen isotopes

I'm using carbonate $\delta^{18}O$ data from 2 Carnian and 4 Anisian sites. The lines on the plot show predicted average continental carbonate values as a function of temperature for the model prior (black) and posteriors (colors as throughout). This is derived in two steps:

* Precipitation $\delta^{18}O$ value is a function of temperature. The prior uses the classic modern empirical relationship. The slope and intercept of the relationship are model parameters which are sampled separately and optimized for each stage.
* Carbonate $\delta^{18}O$ value is a function of the local precipitation $\delta^{18}O$ value and temperature. Previously we had been using MAT to describe fractionation in this step, which is reflected in the solid lines below. Now we're using summer season temperature. Because the relationship between MAT and summer temperature is a function of both MAT and latitude we can't plot a single line representing this, but I've approximated the relationship by plotting the predicted carbonate $\delta^{18}O$ values using summer season temperature for a site at 45$^\circ$ latitude (dotted lines).

The data are plotted as above, with the posterior median MAT on the x axis and the measured proxy value on the y axis. 

```{r oxygen, echo=FALSE}
mat = 0:40
c_cont = -mat + 1.4836 - 0.2738 * 45
t_seas = (-0.9189 + sqrt(0.9189 ^ 2 - 4 * -0.0015 * c_cont)) /
      (2 * -0.0015)
d18O.prior = -14 + 0.59 * mat
d18Oc.prior = (d18O.prior + (2.78e6 / (mat + 273) ^ 2 + -2.98)) * 0.97002 - 29.98
d18Oc.prior.seas = (d18O.prior + (2.78e6 / (t_seas + 273) ^ 2 + -2.98)) * 0.97002 - 29.98

d18O.car.post = c(pm$d.car) + c(pm$e.car) * mat
d18Oc.car.post = (d18O.car.post + (2.78e6 / (mat + 273) ^ 2 + -2.98)) * 0.97002 - 29.98
d18Oc.car.post.seas = (d18O.car.post + (2.78e6 / (t_seas + 273) ^ 2 + -2.98)) * 0.97002 - 29.98

d18O.an.post = c(pm$d.an) + c(pm$e.an) * mat
d18Oc.an.post = (d18O.an.post + (2.78e6 / (mat + 273) ^ 2 + -2.98)) * 0.97002 - 29.98
d18Oc.an.post.seas = (d18O.an.post + (2.78e6 / (t_seas + 273) ^ 2 + -2.98)) * 0.97002 - 29.98

plot(mat, d18Oc.prior, type = "l", 
     ylim = range(c(d18Oc.prior, d18Oc.an.post, d18Oc.car.post,
                    pc_samples.car$d18O, pc_samples.an$d18O)),
     xlab = "MAT", ylab = expression(delta^{18}*"O terrestrial carb"))
lines(mat, d18Oc.prior.seas, lty = 3)
lines(mat, d18Oc.car.post, col = col.c)
lines(mat, d18Oc.car.post.seas, col = col.c, lty = 3)
lines(mat, d18Oc.an.post, col = col.a)
lines(mat, d18Oc.an.post.seas, col = col.a, lty = 3)

points(pm$t_cont.car[pc_sites.ind.car], pc_samples.car$d18O, pch = 20, col = col.c)
points(pm$t_cont.an[pc_sites.ind.an], pc_samples.an$d18O, pch = 20, col = col.a)

text(pm$t_cont.car[pc_sites.ind.car], par("usr")[4], pc_sites.ind.car, 
     xpd = NA, pos = 3, col = col.c)

text(pm$t_cont.an[pc_sites.ind.an], par("usr")[4], pc_sites.ind.an, 
     xpd = NA, pos = 3, offset = 1.25, col = col.a)
```

The big take-home here is that terrestrial carbonate $\delta^{18}O$ values are low relative to the prior, especially at the warmer sites. Most of the values would be consistent with MATs below freezing. Accounting for seasonality reduces this effect, because mineral precipitation at warmer summer temperatures produces calcite with lower $\delta^{18}O$. Even then, however, some of the data (especially for Anisian sub-tropical sites, e.g., sites 2 and 6) would imply pretty low MAT (near or just below zero).

The model accommodates this by changing the relationship between MAT and precipitation $\delta^{18}O$ values for the Anisian...precipitation $\delta^{18}O$ at any given MAT is 1-2 per mil lower in the Anisian than the modern or the Carnian (where the data are more consistent w/ the modern precipitation isotope gradient, after accounting for temperature seasonality) 

## Summary

Univariate interpretations of the terrestrial proxy data at all sites *except* Utah suggest MATs no higher than 20$^\circ$. Adding variable $\delta^{18}O$ of seawater and the data from the Antarctic site has pulled the Anisian continental temperatures at all latitudes up, so that in most cases they are at the higher end of the ranges suggested by the terrestrial proxy data. Temperatures for the Carnian tend to remain a compromises, with Al/Si values pulling these temperatures warmer and CIA values suggesting lower MATs than reflected in the JPI posterior.
